The Evolution of Phenotypically Invariant Genetic Networks

### 2017-06-01 handnotes.tex
handnotes contains a simple  explanantion of the model.

\documentclass[11 pt]{article}
\usepackage{amsmath, amssymb, color, xcolor}
\usepackage{graphicx, wrapfig, float, caption, dsfont}
\usepackage{fullpage}
\usepackage[backref=page, hidelinks, colorlinks=true, citecolor=blue!60!black!100]{hyperref}
\usepackage{tikz}
\usetikzlibrary{arrows.meta, shapes}
\usepackage{caption, subcaption}
\usepackage{natbib} % gives us \citet: Author (year) and \citep: (Author; year)
\usepackage{authblk}

@@

@@

\paragraph{To do, soon:}

\begin{enumerate}

  \item \emph{(josh)} Add biology to introduction. (keep an eye out for concrete numbers we can use!) \jss{added some relevant biology citations to introduction. Still needs a lot of refinement.} \ref{sec:intro}

    \item Do some order-of-magnitude calculation to figure out how fast systems should drift.
        \begin{itemize}
            \item \emph{(peter)} Work out the math for how to do this.
            \item \emph{(josh)} Find real parameters to put into this. 
              \jss{added some notes for TFBS gain and loss calculations from Lynch} \ref{sec:param}
        \end{itemize}

      \item \emph{(josh)} Write out simple examples in biological words. \ref{sec:examples}

    \item \emph{(josh)} Rephrase hybrid inviability example in biological terms.  Discuss F1's (or heterozygotes) versus F2's.

    \item \emph{(josh)} Clarify our assumptions around $x(0)$.

    \item \emph{(josh)} Note we don't include $D$ for simplicity but don't.

    \item \emph{(peter)} Figure out how to specify $P$ for Kalman decomposition.

    \item \emph{(josh \jss{\checkmark},peter)} Prove or find citation for uniqueness of minimal system up to basis change. \jss{See \citep[Chapter 10.6, Theorem 6.9: Fundamental Theorem of Linear Realization Theory.]{kalman1969topics}}

    \item Work examples (minimal; not minimal; with/without B,C fixed) of Kalman decomposition and description of moving along the walk.
        \begin{itemize}
            \item \emph{(josh)} Work out parameterization.
            \item \emph{(peter)} Describe Brownian motion.
        \end{itemize}

    \item \emph{(peter)} Write down and decide between options for parameterizing the random walk on the neutral manifold:
        use Kalman decomposition, and parameterize Brownian motion so that it has constant speed;
        OR consider Josh's parameteriziation like this;
        OR just say we do Brownian motion on the manifold (more or less implicit characterization).

    \item \emph{(peter)} Describe model leading to constraints on regulation so we have a stationary distribution.
        Ask questions of stationary distribution: robustness, connectivity, xxx?

    \item Consider options for fitness function: Euclidean? Sign of eigenvalues?  Time domain or frequency?
      First try on simple concrete examples. \jss{see \ref{sec:del}}

    \item Figure out coherent model for changing dimension: do coefficients stay stuck at zero? In what parameterization? 
        Do we need to allow deleterious variation to get deletion?

    \item Write something in discussion about nonlinear systems.  Do an example if there is a feasible one.

    \item Simulate.

\end{enumerate}

% Outline:
% 
% - What the problems are:
% 
%     * evolution of perfectly fit populations
% 
% - Motivations: Can you infer function from structure? And...
% 
%     * motifs
%     * speciation
%     * network size
%     * rube-goldberginess
%     * redundancy, robustness
%     * evolvability related to minimality?
% 
% - Give a concrete, simple example.
% 
% - Modeling framework (LTI systems)
%
%     * y = "phenotype"
%     * x = "internal state" or "kryptotype"?
%     * A = "regulatory matrix" or "genetic network"?
%     * Kalman decomposition
% 
% - Evolution of free parameters in the Kalman decomposition
%
%     * Give or refer to earlier simple description of population dynamics which makes the system evolve.
%     * What do we want to know from this model:
%
%       - How fast do the internal dynamics change?
%       - Stationary distribution on network complexity/shape/properties
%       - How fast does incompatibility accumulate?
%
%     * Brownian motion on free parameters
%     * how mutations map into this space
%     * dimension: duplication, deletion, recruitment of new things
% 
% - (Compute things)
%
% - Generalize ideas to nonlinear systems
%
%   * compute isomorphic or dynamically equivallent nonlinear systems of the feed-forward motif or of the three-node adaptation motif. 
% 
% - (Allow non-fit stuff too.)

%%% Questions:

% is y the phenotype or H(z) the phenotype? Do we need new jargon anyway?

\title{The Evolution of Phenotypically Invariant Genetic Networks}
\author[1]{Joshua S. Schiffman}
\author[1,2]{Peter L. Ralph}
\affil[1]{University of Southern California, Los Angeles, California}
\affil[2]{University of Oregon, Eugene, Oregon}
\affil[ ]{\texttt{jsschiff@usc.edu} \qquad \texttt{plr@uoregon.edu}}
\maketitle

\begin{abstract}
I will outline an analytical theory to study the evolution of biological systems such as gene regulatory networks, borrowing insight and tools from control engineering, systems identification, and dynamical systems theory. I will describe a null model of regulatory network evolution by analytically describing the set of all linear gene networks (of any size) that produce identical phenotypes -- and the evolutionary paths connecting them. In the idealized case of a perfectly adapted population, constant selection, and a static environment, we observe neutral evolution as a random walk over the phenotypically- invariant network-space. Under neutral conditions, this model can provide descriptions of expected network size and connectivity under mutation-selection equilibrium, estimate the rate of regulatory rewiring, and the rates at which Dobzhansky-Muller incompatibilities arise in reproductively isolated populations. This analysis provides insight into the mechanisms and parameters important for understanding developmental systems drift, network rewiring, evolvability, epistasis, and speciation, as well as the tenuous connection between network architecture and function. 
\end{abstract}

\section{Background} \label{sec:intro}

\plr{Review of relevant other stuff.}

\jss{Introduction}

@@

\plr{phenomenologically or dynamically equivalent?}
The set of all systems -- that is all model structures and all parameter schemes -- that
are externally, or phenomenologically, equivalent, biologically realizable, and mutationally accessible,

What are typically referred to as ``design principles'' or some sort of clear structure-function relationships are commonly saught \citep{milo2002network, alon2007network, ma2009defining}.

Seeking to attribute functionality, uninformed by evolution, can lead to spurious claims
(i.e. see Graur \citep{lynch2007evolution, lynch2007frailty, graur2013immortality}).

A linear time invariant (LTI) dynamical system is a system of linear
differential equations that describes a physical process. 
LTI systems are used in electrical and control engineering to model a myriad of phenomena including
circuits. 
One can use this methodology, under a set of assumptions, to reverse engineer a
mechanism from impulse experiments (input/output data). The idea is that given
a black box, and it's experimental manipulation, can we describe the internal
mechanisms?

\section{The problem}

Organisms often have to respond to their environments,
and the way this happens is that external input triggers a cascade of molecular signals
whose eventual result is a reaction,
whose approprateness for the situation determines fitness.
\plr{A similar situation occurs in development (...?).}
However, there are many ways to skin a cat \jss{(there are also many ways to develop a cat)}:
especially given a number of possible molecular intermediaries,
there may be a large number of ways to construct such a molecular signalling system
that has precisely the \emph{same} input-output relationships,
and hence the same function.
Therefore, any mutation that changes one such system to another, equivalent version
will be neutral,
and may well become polymorphic in a population if there are no ill effects in heterozygotes.
Certainly, mutations that move a system away from optimal functioning will also appear,
but in very large populations will not rise to high frequency.
\plr{Deleterious things are quite important but somehow justify starting with the neutral ones.}

\plr{
    More stuff on why we care about this:  
Provide neutral model; can address questions.  
Main point of paper.
}

\begin{example}[Example: Oscillator.] \label{ex:osc}
Motivate the system:
\begin{align*}
    A &= \left[\begin{matrix} 
        0 & 1 \\ 
       -1 & 0 
    \end{matrix}\right]
\end{align*}

    \begin{figure}
        \begin{center}
            \begin{tikzpicture}
            \begin{scope}[every node/.style={circle,thick,draw}]
                \node (A) at (0,0) {$x_{1}$};
                \node[dashed] (B) at (3,0) {$x_{2}$};
                \node[shape=rectangle] (U) at (1.5,2) {input};
                \node[shape=rectangle] (y) at (1.5,-2) {output};
            \end{scope}

            \begin{scope}[>={Stealth[black]},
                          every node/.style={fill=white,circle},
                          every edge/.style={draw=black, thick}]
                \path [->, >=Rectangle] (A) edge[bend left] node {\tiny $-1$} (B);
                \path [->] (B) edge[bend left] node {\tiny $1$} (A); 
                \path[->] (U) edge[dashed] node {\tiny $1$} (A);
                \path[->] (U) edge[dashed] node {\tiny $1$} (B);
                \path[->] (A) edge[dashed,bend right] node {\tiny $1$} (y);
            \end{scope}
            \begin{scope}[>={Stealth[black]},
                          every edge/.style={draw=black, thick}]
                %\path [->] (A) edge[loop left] node {\tiny $\lambda_{1}$} (A);
                %\path [->] (B) edge[loop left] node {\tiny $\lambda_{2}$} (B);
            \end{scope}

            \end{tikzpicture}
        \end{center}
        \caption{
            Diagram of Example \ref{ex:osc} in the text.  
            \plr{explain what arrows mean if nec}
            \label{fig:oscillator_diagram}}
    \end{figure}

as a special case of the below.

\end{example}

\begin{example}[Example: two-component systems.] \label{ex:2x2}
As a more general example,
suppose that a cell produces two molecular species, $S_1$ and $S_2$ 
whose production are both stimulated by the presence of an external ``input'' substance.
\plr{call them TF?}
Suppose that the two species self-regulate with strengths $\lambda_1$ and $\lambda_2$, respectively,
and the second species also regulates the first species with strength $\gamma$.
\plr{Rephrase in terms of binding to promoter sequences?}
However, only the time course of the concentration of
the first one of these species determines the fitness of the organism.
(As happens for instance in \plr{XXX}.)
Write $x(t) = (x_1(t),x_2(t))$ for the concentrations of the two species at time $t$,
and write $u(t)$ for the concentration at time $t$ of the input substance.
Then, if the rates at which each species are produced
are linear functions of the concentrations
\plr{(look up way this is usually said in the literature.)},
then the dynamics of the system are given by
\begin{align*}
    \ddt x_1(t) &= \lambda_1 x_1(t) + \gamma x_2(t) + u(t) \\
    \ddt x_2(t) &= \lambda_2 x_1(t) + u(t) ,
\end{align*}
where $\ddt x$ denotes the time derivative.
The initial conditions, $x_1(0)$ and $x_2(0)$, 
and the input $u(t)$ then determine the concentrations through time.
If we record the regulatory coefficients in the matrix:
\begin{align*}
    A &= \left[\begin{matrix} 
        \lambda_1 & \gamma \\ 
       0 & \lambda_2  
    \end{matrix}\right],
\end{align*}
and define the column vector $B = [1,1]^T$,
then in matrix notation the dynamics are
\begin{align*}
    \ddt x(t) &= A x(t) + B u(t) .
\end{align*}
If these are all transcription factors,
and we suppose their binding motifs are fixed,
then the $i^\text{th}$ row of $A$ is determined by the $i^\text{th}$ promoter sequence.

\end{example}

Since the system is linear, the state of the system at any time
is the superposition of its responses to all previous inputs.
This implies that 
if $G(t)$ is the transient response of the system to a unit impulse,
$t$ units of time later,
and $x(0) = 0$,
then the time course of the concentration of the thing we care about, $x_1(t)$, 
can be written as
\begin{align} \label{eqn:ex_convolution}
    x_1(t) &= \int_0^t G(t-s) u(s) ds .
\end{align}
However, it turns out that 
\plr{look for simple demontration this is true}
\begin{align*}
    P_{p} = \left[\begin{matrix} 
        1 & 0 \\ 
        p & 1-p 
    \end{matrix}\right],
\end{align*}
then for any $p \neq 1$, 
if we replace the regulatory matrix $A$ with
\begin{align*}
    A(p) = P A P^{-1},
\end{align*}
then the response of $x_1(t)$ to any particular input $u(t)$ is identical to the original system.
(However, $x_2$ may well be different!)

For instance, setting $p=-1$,
the system with
\begin{align*}
    A(-1) = \left[\begin{matrix} 
        \lambda_1 + \gamma/2 & \gamma/2 \\ 
        \lambda_2-\lambda_1-\gamma/2 & \lambda_2 - \gamma/2
    \end{matrix}\right],
\end{align*}
looks very different, but gives the same input-output relationship.
Although these systems are equivalent,
hybrids between them may not be:
\plr{do example}.

\begin{example}[Hyrbid Inviability]
  Let, \begin{align*}
    A(0) =\left[\begin{matrix}
      0 & 1 \\
      -1 & 0
    \end{matrix}\right]
    \text{and }
    A(2) = \left[\begin{matrix}
      2 & -1 \\
      5 & -2
    \end{matrix}\right]
  \end{align*}
  from Example \ref{ex:osc} ($B=\begin{bmatrix} 1 \\ 1 \end{bmatrix}$ and $C=\begin{bmatrix} 1 & 0 \end{bmatrix}$). 
If hybrids are formed by exchanging genes (swapping rows), the hybrid's gene network is,
  \begin{align*}
    \text{hybrid network } := \mathcal{H}(A(0), A(2)) = \left[\begin{matrix}
      2 & -1 \\
      -1 & 0
    \end{matrix}\right] ,
  \end{align*}
  \jss{ Or should it be: \begin{align*}
    \text{hybrid network } := \text{mean}(A(0), A(2)) = \left[\begin{matrix}
      1 & 0 \\
      2 & -1
    \end{matrix}\right] ,
  \end{align*} }
  It's transfer function will be $H(z, \mathcal{H}) = \frac{z-2}{z^2 - 3z +1}$, which produces very different dynamics than its parents, $H(z, A(0)) = H(z, A(2)) = \frac{z+1}{z^2 + 1}$. The time dynamcs of molecular species $1$ in the hybrid is $y^{(\mathcal{H})} =e^{t} \cosh(\sqrt{2} t)$ \jss{ which is approx. $\approx e^{t}$}, with the concentration of species $1$ rapidly heading towards infinity. The dynamics of species $1$ in the parental networks is $y = \sin(t) + \cos(t)$, where it continuously oscillates.

  \jss{The transfer function $H(z, \mathcal{H})$, by averaging networks, is $\frac{1}{z-1}$ and the time dynamics are: $y^{(\mathcal{H})} = e^{t}$.}
\end{example}

\begin{definition}[Biological Species Concept]
  Let $\mathcal{A}$ be the set of all phenotypically symmetric gene networks with transfer funtion $H(z)$. 
  Choose $A$ and $\widetilde A$, both in $\mathcal{A}$. 
  If $\mathcal{H}(A, \widetilde A) \notin \mathcal{A}$ then $A$ and $\widetilde A$ are distinct species. 
\end{definition}

\begin{figure}
  \centering
  \begin{subfigure}{0.5\textwidth}
    \centering
\begin{tikzpicture}
\begin{scope}[every node/.style={circle,thick,draw}]
  \node (A) at (0,0) {$x_{1}$};
    \node[dashed] (B) at (4,0) {$x_{2}$};
    \node[shape=rectangle] (U) at (2,2) {input};
    \node[shape=rectangle] (y) at (2,-2) {output};
\end{scope}

\begin{scope}[>={Stealth[black]},
              every node/.style={fill=white,circle},
              every edge/.style={draw=black, thick}]
    %\path [->] (A) edge[bend left] node {$c$} (B);
    \path [->] (B) edge node {\tiny $\gamma$} (A); 
    \path[->] (U) edge[dashed] node {\tiny $1$} (A);
    \path[->] (U) edge[dashed] node {\tiny $1$} (B);
    \path[->] (A) edge[dashed, bend right] node {\tiny $1$} (y);
\end{scope}
\begin{scope}[>={Stealth[black]},
              every edge/.style={draw=black, thick}]
    \path [->] (A) edge[loop left] node {\tiny $\lambda_{1}$} (A);
    \path [->] (B) edge[loop right] node {\tiny $\lambda_{2}$} (B);
\end{scope}

\end{tikzpicture}
    \caption{Gene Regulatory Network $A(0)$.}
  \end{subfigure}%
  \begin{subfigure}{0.5\textwidth}
    \centering
\begin{tikzpicture}
\begin{scope}[every node/.style={circle,thick,draw}]
  \node (A) at (0,0) {$x_{1}$};
    \node[dashed] (B) at (4,0) {$x_{2}$};
    \node[shape=rectangle] (U) at (2,2) {input};
    \node[shape=rectangle] (y) at (2,-2) {output};
\end{scope}

\begin{scope}[>={Stealth[black]},
              every node/.style={fill=white,circle},
              every edge/.style={draw=black, thick}]
    \path [->, sloped] (A) edge[bend left] node {\tiny $(\lambda_{2} - \lambda_{1} - \frac{\gamma}{2})$} (B);
    \path [->, sloped] (B) edge[bend left] node {\tiny $\frac{\gamma}{2}$} (A); 
    \path[->] (U) edge[dashed] node {\tiny $1$} (A);
    \path[->] (U) edge[dashed] node {\tiny $1$} (B);
    \path[->] (A) edge[dashed, bend right] node {\tiny $1$} (y);
\end{scope}
\begin{scope}[>={Stealth[black]},
              every edge/.style={draw=black, thick}]
    \path [->] (A) edge[loop left] node {\tiny $(\lambda_{1} + \frac{\gamma}{2})$} (A);
    \path [->] (B) edge[loop right] node {\tiny $(\lambda_{2}-\frac{\gamma}{2})$} (B);
\end{scope}

\end{tikzpicture}
    \caption{Gene Regulatory Network $A(-1)$.}
  \end{subfigure}
  \caption{Neutral rewiring of gene network $A(p)$ from Example \ref{ex:2x2}. Not only are the regulatory coefficients different between $A(0)$ and $A(-1)$, there is also a new regulatory connection.}
\end{figure}

\section{Regulatory networks as linear, time-invariant systems}

\plr{Make sure to have strongly worded disclaimer somewhere
that we know real networks aren't linear
but they are locally.
Plus, it works in engineering.}

\jss{Maybe we should discuss the Hartman-Grobman Theorem?}

% - General description of the LTI process in biological context
%     * y = "phenotype"
%     * x = "internal state" or "kryptotype"?
%     * A = "regulatory matrix" or "genetic network"?
% - Why transfer function determines input-output relationship
% - Kalman decomposition in unique form

We will now lay out the model in more general terms.
Suppose that the \emph{internal state} of the system
is parameterized by the concentrations of a collection of $n$ molecular species,
$S_1, \ldots, S_n$,
and the vector of concentrations at time $t$ we denote $x(t)=(x_1(t),\ldots,x_n(t))$.
There are also $m$ ``input'' species, whose concentrations are determined
exogenously to the system,
and are denoted $u(t) = (u_1(t),\ldots,u_m(t))$,
and $\ell$ ``output'' species, whose concentrations are denoted
$y(t) = (y_1(t),\ldots,y_\ell(t))$.
The output is merely a linear function of the internal state:
\begin{align*}
    y_i(t) = \sum_j C_{ij} x_i(t).
\end{align*}
Since $y$ is what natural selection acts on, we refer to it as the \emph{phenotype},
and sometimes in contrast refer to $x$ as the \emph{kryptotype},
as it is ``hidden'' from direct selection.
\plr{Maybe.}
The rate at which the $i^\text{th}$ species is produced
is a weighted sum of the concentrations of the other species
as well as the input:
\begin{align*}
    \ddt x_i(t) = \sum_j A_{ij} x_j(t) + \sum_k B_{ik} u_k(t) .
\end{align*}
In matrix notation, this is written more concisely as
\begin{align} \label{eqn:lti_system}
    \ddt x(t) &= A x(t) + B u(t) \\
    y(t) &= C x(t) .
\end{align}

\plr{Everything we say I think is assuming $x(0)=0)$.  
Figure this out. ``Zero-state equivalence.''}

Given an initial condition and an input, 
it is possible to write the solution $x(t)$
as a convolution with an input-response kernel as in the example above
(equation \eqref{eqn:ex_convolution}).
An alternative way of describing the input-output relationship,
more common in engineering,
is instead to find the \emph{transfer function}
\begin{align} \label{eqn:transfer_fn}
    H(z) = C(z I - A)^{-1} B,
\end{align}
which is more mathematically readable \plr{if you know what it means}.
This can be thought of as the system's response to input at ``frequency'' $z$.
The fact that the transfer function uniquely determines the system's
input-output relationship (i.e., the mapping $u \mapsto y$)
follows from the fact that if $\tilde y(z) = \int_0^\infty e^{-zt} y(t) dt$
is the Laplace transform of $y$,
and $\tilde u(z)$ is the Laplace transform of $u$,
then
\begin{align*}
    \tilde y(z) = H(z) \tilde u(z) .
\end{align*}
(This happens because the Laplace transform takes convolutions to products,
and $H$ is the Laplace transform of the kernel $G$.)
Concretely,

\plr{Should we remind what dimensions everything is?}

The state space representation is considered an internal description of the
system, whereas the transfer function is an external description. 
``Realizations'' are something.

\begin{definition}[Phenomenological equivalence of systems]
    Let $(x(t),y(t))$ and $(\bar x(t),\bar y(t))$ be the solutions to \eqref{eqn:lti_system}
    with coefficient matrices $(A,B,C)$ and $(\bar A,\bar B,\bar C)$ respectively,
    and both $x(0)$ and $\bar x(0)$ are zero. 
    The systems defined by $(A,B,C)$ and $(\bar A,\bar B,\bar C)$ are
    \textbf{phenomenologically equivalent} 
    if
    \begin{align*}
        y(t) = \bar y(t) \qquad \text{for all} \; t \ge 0.
    \end{align*}
    Equivalently, this occurs if and only if
    \begin{align*}
        H(z) = \bar H(z)  \qquad \text{for all} \; z \ge 0,
    \end{align*}
    where $H$ and $\bar H$ are the transfer functions of the two systems.
\end{definition}

One way to find other systems equvalent to a given one
is by change of coordinates (``algebraic equivalance''):
if $T$ is an invertible matrix, then the systems $(A,B,C)$ and $(TAT^{-1},TB,CT^{-1})$
have the same dyanamics because their transfer functions are equal:
\begin{align*}
    CT^{-1}( zI - TAT^{-1})^{-1}TB
    =
    CT^{-1}T( zI - A)^{-1}T^{-1}TB
    =
    C( zI - A)^{-1}B .
\end{align*}
However, the converse is not necessarily true: 
systems can have identical transfer functions without being changes of coordinates of each other.
In fact, systems with identical transfer functions can involve interactions between different
numbers of molecular species.

The set of all systems phenomenologically equivalent to a given system $(A,B,C)$ 
is elegantly described using the Kalman decomposition,
which also clarifies the system dynamics? tells us a lot about how it works? \plr{or something}
To motivate this, first note that the the input $u(t)$ only directly pushes the system
in directions lying in the span of the columns of $B$.
As a result, different combinations of input can 
move the system in any direction that lies in the \emph{reachable subspace},
which we denote by $\reachable$,
and is defined to be the closure of $\spn(B)$ under applying $A$
(or equivalently, the span of $B, AB, A^2B, \ldots A^{n-1}B$).
Analogously to this, we define
the \emph{observable subspace}, $\mathcal{O}$,
to be the closure of $\spn(C^T)$ under applying $A$.
(Or: $\unobservable$ is the largest $A$-invariant subspace
contained in the null space of $C$;
and $\reachable$ is the largest $A$-invariant subspace contained in the image of $B$.)

If we define
\begin{enumerate}
    \item The columns of $P_\rno$ are an orthonormal basis for $\reachable \cap \unobservable$.
    \item The columns of $P_\ro$ are an orthonormal basis of
        the complement of $\reachable \cap \unobservable$ in $\reachable$.
    \item The columns of $P_\nro$ are an orthonormal basis of
        the complement of $\reachable \cap \unobservable$ in $\unobservable$.
    \item The columns of $P_\nrno$ are an orthonormal basis of
        the the remainder of $\R^n$.
\end{enumerate}
If we then define
\begin{align*}
    P &= 
    \left[ \begin{array}{c|c|c|c}
        P_\rno & P_\ro & P_\nro & P_\nrno
    \end{array} \right] ,
\end{align*}
then
\begin{align*}
    P^T P
    &=
    \left[ \begin{array}{c|c|c|c}
        I & 0 & 0 & 0 \\
        \hline
        0 & I & U & 0 \\
        \hline
        0 & V & I & 0 \\
        \hline
        0 & 0 & 0 & I 
    \end{array} \right] .
\end{align*}
\plr{Check this.  Can we get $U=V=0$?}

The following theorem can be found in SOME REFERENCE.

\begin{theorem}[Kalman decomposition] \label{thm:kalman}
        For any system $(A,B,C)$ with corresponding Kalman basis matrix $P$,
        the transformed system $(PAP^{-1},PB,CP^{-1})$  has the following form:
        \begin{align*}
            \widehat A = PAP^{-1}
            &=
            \left[ \begin{array}{cccc}
                A_{\rno} & A_{\rno,\ro} & A_{\rno,\nrno} & A_{\rno,\nro} \\
                0 & A_{\ro} & 0 & A_{\ro,\nro} \\
                0 & 0 & A_{\nrno} & A_{\nrno,\nro} \\
                0 & 0 & 0 & A_{\nro}
            \end{array} \right] ,
        \end{align*}
        and
        \begin{align*}
            \widehat B = PB
            &=
            \left[ \begin{array}{cccc}
                B_{\rno} \\
                B_{\ro} \\
                0 \\
                0 
            \end{array} \right] ,
        \end{align*}
        and
        \begin{align*}
            \widehat C = CP^{-1}
            &=
            \left[ \begin{array}{cccc}
                0 & C_{\ro} & C_{\nrno} & 0 
            \end{array} \right] .
        \end{align*}
        The transfer function of both systems is given by
        \begin{align*}
            H(z) = C_{\ro} ( zI - A_{\ro} )^{-1} B_{\ro} .
        \end{align*}
\end{theorem}

In the latter case, we say that the system is \emph{minimal} 
-- there is no equivalent system with a smaller number of species.
Note that this says that any two equivalent minimal systems
are changes of basis of each other.

Since any system can be put into this form,
and once in this form, its transfer function is determined only by 
$C_{\ro}$, $A_{\ro}$, and $B_\ro$,
therefore, the set of all equivalent systems are parameterized by
the dimension $n$,
the choice of basis ($P$),
the remaining submatrices in $\widehat A$, $\widehat B$, and $\widehat C$
(which are unconstrained),
and a invertible transformation of $\spn(P_{\ro})$, which we call $T_\ro$.

\begin{theorem}[Parameterization of equivalent systems]
    Let $(A,B,C)$ be a minimal system.
    \begin{itemize}
        \item[(a)]
            Every equivalent system is of the form given in Theorem \ref{thm:kalman},
            i.e., can be specified by choosing a dimension, $n$;
            submatrices in $\widehat A$, $\widehat B$, and $\widehat C$ 
            except for $A_\ro=A$, $B_\ro=B$, and $C_\ro=C$;
            and choosing an invertible matrix $P$.

        \item[(b)]
            \plr{conjecture:}
            The parameterization is unique
            if $P$ is futhermore chosen so that 
            each $P_x$ other than $P_\ro$ is a projection matrix,
            and that 
            \begin{align*}
                0
                =
                P_x^T P_y
            \end{align*}
            for all $(x,y)$ except $(\ro,\nrno)$.

    \end{itemize} 
\end{theorem}

\plr{Another way of saying it: pick the $\reachable$ and $\unobservable$ subspaces,
that must intersect in something of the minimal dimension;
then let $P$ be the appropriate basis?}

In some situations we may be interested in only ``network rewiring'',
where $A$ changes while $B$ and $C$ do not.
For instance, 
if all non-regulatory functions of each molecule are strongly constrained,
then $C$ cannot change.
Likewise, if responses of each molecule to the external inputs are not changed by evolution,
then $B$ does not change.

\begin{lemma}
    Let $(A,B,C)$ be a system.
    Every matrix $\widehat A$ such that $(\widehat A,B,C)$ is equivalent to $(A,B,C)$
    can be specified in the form given in Theorem \ref{thm:kalman},
    with \plr{them having the same $\ro$ bit}
    and satisfying
    \begin{align*}
        \widehat B &= P B \\
        \widehat C P &= C
    \end{align*}
\end{lemma}

It is remarkable to note that even with the relationships 
between environment, cryptotype, and phenotype constant,
and in the minimal dimension,
there are still almost always degrees of freedom.
These correspond to distinct genetic networks
that perform indistinguishable functions.
For example, the equivalent systems of Example \ref{ex:2x2} above
are minimal, and share common $B$ and $C$ matrices.

\plr{Maybe say that stability to noise can be determined by $A_\nro$ (and others?),
while complexity of the internal, unobserved dynamics
are somehow given by the difference between the dimension of $y$ and the dimension of $A_\ro$.}

\plr{General argument why Brownian motion on set of equivalent things is maybe ok.}

>>>>

\documentclass[9 pt]{article}
\usepackage{amsmath, amssymb, color, xcolor}
\usepackage{graphicx, wrapfig, float, caption, dsfont, bbm}
\usepackage{fullpage}
\usepackage[backref=page, hidelinks, colorlinks=true, citecolor=blue!60!black!100]{hyperref}
\usepackage{tikz}
\usetikzlibrary{arrows.meta, shapes}
\usepackage{caption, subcaption}
\usepackage{natbib} % gives us \citet: Author (year) and \citep: (Author; year)
\usepackage{authblk}

\usepackage{multicol}

@@

\newcommand{\A}{\mathcal{A}}
\newcommand{\diag}{\text{diag}}
\newcommand{\1}{\mathbbm{1}}

@@

\section*{The Evolution of Phenotypically Invariant Genetic Networks}
Joshua S. Schiffman and Peter L. Ralph

\begin{multicols}{2}

\section*{Abstract}
I will outline an analytical theory to study the evolution of biological systems such as gene regulatory networks, borrowing insight and tools from control engineering, systems identification, and dynamical systems theory. I will describe a null model of regulatory network evolution by analytically describing the set of all linear gene networks (of any size) that produce identical phenotypes -- and the evolutionary paths connecting them. In the idealized case of a perfectly adapted population, constant selection, and a static environment, we observe neutral evolution as a random walk over the phenotypically- invariant network-space. Under neutral conditions, this model can provide descriptions of expected network size and connectivity under mutation-selection equilibrium, estimate the rate of regulatory rewiring, and the rates at which Dobzhansky-Muller incompatibilities arise in reproductively isolated populations. This analysis provides insight into the mechanisms and parameters important for understanding developmental systems drift, network rewiring, evolvability, epistasis, and speciation, as well as the tenuous connection between network architecture and function.

\section*{Introduction}

@@

\section*{The Model I: Gene Networks as Linear Systems}

  Organisms' phenotypes are constructred by complex gene by gene and gene by environment interactions. Although a very simplified description, here we define the phenotype to be the spatio-temporal molecular dynamics directly under natural selection. Basically the \emph{what}, \emph{when}, and \emph{where}, of an organism's molecules that are physiologically or otherwise important and relevant to survival. Thus we say that some function $f(t)$ is a phenotype where, 
  \begin{align}
    f(t) = \int_{0}^{\infty} h(t) u(t) dt  ,
  \end{align}
  and $h(t)$ is the \emph{impulse response} or \emph{kernel} of the system and $u(t)$ is the \emph{input} function. The input can be interpreted as the environment, or as initial conditions, or otherwise, depending on the biological specifics under study.

  Essentially the phenotype $f(t)$ is a consequence of an organism's specific gene by gene interactions, given by $h(t)$, reacting further with the local environment, given by $u(t)$.

  We describe the impulse response as, 
  \begin{align}
    h(t) = C e^{A t} B
  \end{align}
  where $A$ is a gene network -- a square matrix, and $B$ filters and translates the input to the system. The form of $B$ determines precisely how the state of the external environment influences the internal gene network. $C$ filters and translates the dynamics of the system and precesily determines the output, that is, what is visible to selection.

  Generally $A$ can be any real $n \times n$ matrix, $B$ any $n \times \ell$, and $C$ any $\ell \times n$ dimensional matrix. However, for simplicity in exposition (and without loss of generality) we set $\ell = 1$, so that $B$ and $C$ are simply vectors of length $n$.

  Above, $t$ refers to time. Although $f(t)$ describes the phenotype given an input, $h(t)$ describes the phenotype subject only to an impulse -- an input present initially and absent immediately thereafter. Typically, a system $\Sigma$ is defines as,
  \begin{align}
    \Sigma = \left\{ \begin{array}{ll} \dot{x}(t) &= A x(t) + B u(t) \\ y(t) &= C x(t) \end{array} \right.
  \end{align}
  Variables have the same identities as described above and $x(t)$ is a vector of molecule concentrations at time $t$. Therefore the molecular concentrations at a specific time are completely determined by the environmental input and gene by gene interactions. Lastly, a portion and/or combination of these molecules, $y(t)$, are ``observed'' by selection.

  \section*{The Model II: Linear Evolutionary Systems}

  Systems with identical external dynamics do not necessarily have identical internal dynamics. Any linear and minimal system -- minimal, informally meaning that the system's external dynamics are achieved with the fewest possible number of internal components -- has identical external dynamics up to a change of coordinates. 
  \begin{align}
    h(t) &= C e^{A t} B \\
    &= C V^{-1} e^{V A V^{-1} t} V B \\
    &= C V^{-1} V e^{A t} V^{-1} V B \\
    &= C e^{A t} B
  \end{align}
  Two systems, $\Sigma = \{ A, B, C \}$, and $\bar{\Sigma} = \{\bar{A} = VAV^{-1}, \bar{B} = VB, \bar{C} = CV^{-1} \}$, have the same dynamics if they are related by a change of coordinates.

  Although systems may not be identifiable up to a change of coordinates, at present we are primarily interested in a subset of these systems. That is, systems that not only have equivalent external dynamics, but also equivlanet input relationships and under equivalent selective pressures. Formally, systems related by a change of coordinates that leave $B$ and $C$ invariant:
  \begin{align}
    VB &= B \implies \bar{B} = B \\
    CV &= C \implies \bar{C} = C
  \end{align}
  In other words systems with varying genetic architectures yet identical selection pressures, environment, and phenotype.

  Define $V(\tau)$ as the change of coordinates matrix that preserves $B$ and $C$, with $\tau$ a vector of free parameters. The set of \emph{all} phenotypically invariant (minimal) gene networks is, 
  \begin{align}
    A(\tau) = V(\tau) A(0) V^{-1}(\tau) ,
  \end{align}
  and a Linear Evolutionary System is, 
  \begin{align}
    \Sigma(\tau) = \left\{ \begin{array}{ll} \dot{x}(t) &= A(\tau) x(t) + B u(t) \\ y(t) &= C x(t) \end{array} \right .
  \end{align}

  Evolution thus proceeds as a random walk in phenotypically invariant network space.

  \begin{align}
    A(\tau) \xrightarrow{T} A(\tau + \epsilon)
  \end{align}

  \section*{Phenotypic Invariance}

  \section*{Examples}
    \subsection*{Cell Cycle Control or Circadian Rhytm}

    \subsection*{Metabolic Network}

    \subsection*{Gap Gene Network}

    \section*{Discussion}

\bibliographystyle{plainnat}
%\bibliography{krefs}
\end{multicols}
\end{document}
